Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 introduction.tex
minor rewording, added citation
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@951 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the
gravitational-wave amplitude and frequency increasing until the system
eventually merges soon after the innermost stable circular orbit (\ISCO). If a
neutron star is involved it may become tidally disrupted near the merger.  This
disrupted matter can fuel a bright electromagnetic counterpart in the merger's aftermath~\cite{shibata:2007}.
The coalescence of compact binary systems consisting of neutron stars
(\textsc{ns}) and/or black holes (\textsc{bh}) is the most promising source of
gravitational radiation for Advanced \LIGO, Virgo, \GEO\, and
\LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary coalescence
events are expected to be observed in the advanced detector era later this
decade~\cite{Abadie:2010p10836}. In the era of first-generation
detectors the gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
were observed in order to trigger followup observations by optical telescopes.
The best demonstrated latencies were 30--60 minutes~\cite{LOOCUP}, which was
an important achievement, but far too late to catch prompt electromagnetic emission.

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

>>>>

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the
gravitational-wave amplitude and frequency increasing until the system
eventually merges soon after the innermost stable circular orbit (\ISCO). If a
neutron star is involved it may become tidally disrupted near the merger.  This
disrupted matter can fuel a bright electromagnetic counterpart in the merger's aftermath~\cite{shibata:2007}.
The coalescence of compact binary systems consisting of neutron stars
(\textsc{ns}) and/or black holes (\textsc{bh}) is the most promising source of
gravitational radiation for Advanced \LIGO, Virgo, \GEO\, and
\LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary coalescence
events are expected to be observed in the advanced detector era later this
decade~\cite{Abadie:2010p10836}. In the era of first-generation
detectors the gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
were observed in order to trigger followup observations by optical telescopes.
The best demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was
an important achievement, but far too late to catch prompt electromagnetic emission.

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition to reduce the number of templates\cite{Cannon:2010p10398}
We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.
